Grappa: Scalable Computing on Mass-Market Hardware for Irregular, Low-Locality Applications

### 2012-07-21 doc/asplos13/grappa.tex
a few changes to grappa and implementation sections; add figures for system overview and memory structure

\begin{figure}[t]
\begin{center}
  \includegraphics[width=0.5\columnwidth]{figs/grappa-crappy}
\begin{minipage}{0.95\columnwidth}
  \caption{\label{fig:grappa} Grappa runtime system \vspace{-4ex}\TODO{should we use the canonical Grappa picture here augmented withe the software stack?}}
\end{minipage}
\vspace{-3ex}
\end{center}
\end{figure}

@@

\paragraph{Global memory access}
There are two approaches to {\emph accessing} global memory. When the
programmer expects a computation on shared data to have spatial
locality to exploit, {\em cache} operations may be used. When there is no locality to
exploit, {\em delegate} operations are used. Since these operations
are expected communicate with other nodes and have high latency,
\TODO{the runtime systems might context switch? what do we want to say here?}

@@

Grappa contains two key components that are not directly visible to
the programmer but are key to providing good performance. Both depend
on Grappa's ability to tolerate latency in order to obtain this
performance.

The first is a distributed dynamic load balancing mechanism that uses work stealing to
move tasks from busy nodes to idle nodes. When a core detects that it
has idle execution resources and no new tasks to run, it issues steal
requests over the network. The core continues to execute remaining
active work while the steal requests are in flight. When a steal
request succeeds, it returns with more tasks that can be matched with
execution resources and scheduled.

The second is a network packet scheduling and aggregation system. From
the perspective of the application, all communication between tasks in
Grappa takes place either through global shared memory or remote task
spawns. Both of these requests are ultimately translated into explicit
inter-system-node/core communication requests that are dispatched to
the \gasnet~library. Most of \checkme{our requests} are small, and commodity
networks do not transport small packets efficiently. In order to make
more efficient use of bandwidth, Grappa aggregates packets going to the
same destination until it can form a larger packet that will make more
efficient use of the network. In this way, Grappa trades per-request
latency for application-level throughput. \TODO{Jacob, can you add more details here on how this works?}

>>>>

\begin{figure}[t]
\begin{center}
  \includegraphics[width=0.95\columnwidth]{figs/system-overview}
\begin{minipage}{0.95\columnwidth}
  \caption{\label{fig:grappa} Grappa system overview}
\end{minipage}
\vspace{-3ex}
\end{center}
\end{figure}

@@

\paragraph{Global memory access}
There are two approaches to {\emph accessing} global memory. When the
programmer expects a computation on shared data to have spatial
locality to exploit, {\em cache} operations may be used. When there is
no locality to exploit, {\em delegate} operations are used. Since
these operations are expected to communicate with other nodes and have
high latency, cache and delegate operations interact with the
scheduler; when issuing a long-latency request, they suspend the
requesting task and allow the core to execute other work until their
response arrives.

@@

Grappa contains two components that are not directly visible to
the programmer but are key to providing good performance. Both depend
on Grappa's ability to tolerate latency in order to obtain this
performance.

The first is a distributed dynamic load balancing mechanism that uses work stealing to
move tasks from busy cores to idle cores. When a core detects that it
has idle execution resources and no new tasks to run, it issues steal
requests over the network. The core continues to execute remaining
active work while the steal requests are in flight. When a steal
request succeeds, it returns with more tasks that can be matched with
execution resources and scheduled.

The second is a network packet scheduling and aggregation system. From
the perspective of the application, all communication between tasks in
Grappa takes place either through global shared memory or remote task
spawns. Both of these requests are ultimately translated into explicit
inter-system-node/core communication requests that are dispatched to
the \gasnet~library. In our benchmarks, we found that most of our
packets were small (44 bytes on average for BFS). Commodity network
interfaces do not transport small packets efficiently (see
Section~\ref{sec:bandwidth-versus-packet-size}): the rate at which the
network interface will accept new packets of any size is limited, so
the only way to use a respectable fraction of the bandwidth is to send
larger messages.

Grappa's latency tolerance allows us to leverage the parallelism in
our problem domain to build larger packets. Tasks can generate
requests faster than the network will accept new packets, so instead
of immediately sending every message, Grappa queues messages heading
for the same destination. Once enough messages are queued, Grappa
bundles all the messages into a single packet and sends it to the
destination core. Tasks that require a response can suspend, allowing
their core to do other work while waiting for a reply.

While queuing adds latency to each individual request, it is a much
lighter-weight operation than sending a packet and allows us to
amoritize the cost of sending the packet across multiple tasks, while
also making better use of available network bandwidth. In this way,
Grappa trades per-request latency for application-level
throughput.

### 2012-07-21 doc/asplos13/implementation.tex
a few changes to grappa and implementation sections; add figures for system overview and memory structure

\paragraph{2D global addresses} The first global address space is the {\em
two-dimensional} address space. This uses a traditional PGAS addressing model,
where each address is a tuple of a rank in the job (or global process ID) and
an address in that process. The lower 48 bits of the address hold a virtual
address in the process. The top bit is set to indicates that the reference is
a 2D address. This leaves 15 bits for network endpoint ID, which limits our
scalability to $2^15$ endpoints.

@@

All linear addresses refer to data allocated from a global heap. The
heap metadata is stored on a single node. Currently all heap
operations serialize through this node; while this has been sufficient
for our benchmarks, parallel performance can be achieved through
combining \TODO{ cite MAMA and flat combining }. 2D addresses may
refer to memory allocated from a single processes' heap or from a
task's stack.

@@

All global memory operations are implemented with active
messages. Requesters construct a descriptor holding the current
worker \TODO{reorder this paragraph after defined} ID and space for a response, issue their request active
message to the node given by the request address, and suspend
themselves. When the request arrives at the address' home node, the
operation is performed and a response active message is sent. When the
response arrives at the requesting node, the descriptor is filled in
with the results of the operation and the requesting task is woken.

\paragraph{Delegates} Delegate operations are used for low-locality accesses and synchronization. Their modifications
are always executed at the home core of their address. Each task may
have only one outstanding delegate operation \TODO{this is not an
    inherent limitation, it would be better to discuss in terms of
    fence -BM}. We limit delegate
operations to operate on objects in the 2D address space or objects that
fit in a single block of the linear address space. No context switches
are allowed while the data it being modified. Given these
restrictions, we can ensure that delegate operations for the same address from multiple
requesters are always serialized through a single core in the system,
providing atomic semantics without using atomic operations.

>>>>

\begin{figure}[t]
\begin{center}
  \includegraphics[width=0.95\columnwidth]{figs/memory-structure}
\begin{minipage}{0.95\columnwidth}
  \caption{\label{fig:memory-structure} Grappa memory structure}
\end{minipage}
\vspace{-3ex}
\end{center}
\end{figure}

\paragraph{2D global addresses} The first global address space is the {\em
two-dimensional} address space. This uses a traditional PGAS addressing model,
where each address is a tuple of a rank in the job (or global process ID) and
an address in that process. The lower 48 bits of the address hold a virtual
address in the process. The top bit is set to indicates that the reference is
a 2D address. This leaves 15 bits for network endpoint ID, which limits our
scalability to $2^{15}$ endpoints.

@@

All linear addresses refer to data allocated from a global heap. The
heap metadata is stored on a single node. Currently all heap
operations serialize through this node; while this has been sufficient
for our benchmarks, parallel performance can be achieved through
combining \TODO{ cite MAMA and flat combining}. 2D addresses may refer
to memory allocated from a single processes' heap or from a task's
stack. Figure~\ref{fig:memory-structure} shows how 2D and linear
addresses can refer to other cores' memory.

@@

All global memory operations are implemented with active
messages. Requesters construct a descriptor holding the current task
ID and space for a response, issue their request active message to the
node given by the request address, and suspend themselves. When the
request arrives at the address' home node, the operation is performed
and a response active message is sent. When the response arrives at
the requesting node, the descriptor is filled in with the results of
the operation and the requesting task is woken.

\paragraph{Delegates} Delegate operations are used for low-locality accesses and synchronization. Their modifications
are always executed at the home core of their address, and while
arbitrary memory operations can be delegated, we restrict the use of
delegate operations in three ways to make them more useful for
synchronization. First, we limit each task to one outstanding delegate
operation to avoid the possibility of reordering in the
network. Second, we limit delegate operations to operate on objects in
the 2D address space or objects that fit in a single block of the
linear address space so they can be satisfied with a single network
request. Finally, no context switches are allowed while the data is
being modified. Given these restrictions, we can ensure that delegate
operations for the same address from multiple requesters are always
serialized through a single core in the system, providing atomic
semantics without using atomic operations.
